---
title: Example VE_NAI analysis
output: html_document
---

# Data structure

Consider the simulated data structure below.

```{r, sim-data}
# Simulate example data
set.seed(123)
n <- 10000
vax <- rbinom(n, 1, 0.66)
ftime <- rexp(n, rate = 0.1 - 0.06*vax)
# 0 = censored, 1 = symptomatic event, 2 = asymptomatic event
ftype <- sample(0:2, n, replace = TRUE, prob = c(0.9, 0.05, 0.02))  
symp_ind <- as.numeric(ftype == 1)
asymp_ind <- as.numeric(ftype == 2)
# consider a single covariate
age <- rnorm(n, mean = 50, sd = 10)

data <- data.frame(
	ftime = ftime, ftype = ftype, 
	symp_ind = symp_ind, asymp_ind = asymp_ind, 
	vax = vax, age = age
)

head(data)
```

# Estimating Cox models

To estimate the cumulative incidence of any infection, we will fit two Cox models, one for symptomatic and one for asymptomatic.
This allows us to have different baseline hazards for each event type, which feels appropriate given the different methods of ascertainment for the two types of endpoints.

```{r, fit-coxph}
symp_cox_fit <- survival::coxph(survival::Surv(ftime, ftype == 1) ~ vax + age, data = data)
asymp_cox_fit <- survival::coxph(survival::Surv(ftime, ftype == 2) ~ vax + age, data = data)
```

# Estimating conversion probability

We need to estimate probability of having symptoms given infection. For that we can use logistic regression.

```{r, fit-lr}
symp_lr_fit <- glm(
  symp_ind ~ vax + age, 
  # subset to infected ptcpt only
  data = data[data$ftype > 0, ],
  family = binomial()
)
```


# Estimating VE and other parameters

Given these three models, we can estimate the parameters we are interested in.

```{r}
#' Compute VE for naturally asymptomatic infections and other related parameters
#' 
#' @param symp_cox_fit A \code{coxph} fit for symptomatic infections
#' @param asymp_cox_fit A \code{coxph} fit for asymptomatic infections
#' @param symp_lr_fit A \code{glm} fit for probability of symptoms given infection
#' @param data The \code{data} used to fit the above models. Should have columns \code{ftime}, \code{ftype}, 
#' 						 \code{vax}, and other columns for covariates.
#' @param t0 The time(s) of interest. The code computes cumulative incidence at these prespecified times 
#' 					 and reports VE estimates (and other estimates) at each specified time

ve_from_fits <- function(
	symp_cox_fit,
	asymp_cox_fit,
	symp_lr_fit,
	data,
	t0 = quantile(data$ftime, c(0.2, 0.5, 0.9)),
	...
){
	n <- dim(data)[1]
	symp_chaz <- survival::basehaz(symp_cox_fit)
	asymp_chaz <- survival::basehaz(asymp_cox_fit)
	# if the unique times are not the same, then it may indicate a problem with
	# how the models were fit
	stopifnot(all(unique(symp_chaz$time) == unique(asymp_chaz$time)))

	# find times closest to requested t0
	closest_indices <- sapply(t0, function(t) {
  	which.min(abs(symp_chaz$time - t))
	})
	subset_symp_chaz <- symp_chaz[closest_indices,]
	subset_asymp_chaz <- asymp_chaz[closest_indices,]

	surv_data <- merge(subset_symp_chaz, subset_asymp_chaz, by = "time", suffixes = c("_symp", "_asymp"))

	surv_data$total_chaz <- surv_data$hazard_symp + surv_data$hazard_asymp
	# total baseline cumulative incidence of any infection
	surv_data$total_ci <- 1 - exp(-surv_data$total_chaz)

	# predict from models under vaccine/no vaccine
	data0 <- data; data0$vax <- 0
	lp_symp_vax0 <- predict(symp_cox_fit, newdata = data0, type = "lp")
	lp_asymp_vax0 <- predict(asymp_cox_fit, newdata = data0, type = "lp")
	p_symp__inf_vax0 <- predict(
  	symp_lr_fit, newdata = data0, type = "response"
  )

	data1 <- data; data1$vax <- 1
	lp_symp_vax1 <- predict(symp_cox_fit, newdata = data1, type = "lp")
	lp_asymp_vax1 <- predict(asymp_cox_fit, newdata = data1, type = "lp")
	p_symp__inf_vax1 <- predict(
  	symp_lr_fit, newdata = data1, type = "response"
  )

	# conditional cumulative incidence any infection under no vaccine/vaccine
	ci_inf_vax0 <- mapply(
		hazard_symp = surv_data$hazard_symp,
		hazard_asymp = surv_data$hazard_asymp,
		FUN = function(hazard_symp, hazard_asymp){
			1 - exp(- ( hazard_symp * exp(lp_symp_vax0) + hazard_asymp * exp(lp_asymp_vax0) ))
		}
	)
	ci_inf_vax1 <- mapply(
		hazard_symp = surv_data$hazard_symp,
		hazard_asymp = surv_data$hazard_asymp,
		FUN = function(hazard_symp, hazard_asymp){
			1 - exp(- ( hazard_symp * exp(lp_symp_vax1) + hazard_asymp * exp(lp_asymp_vax1) ))
		}
	)
  
  # marginal parameters any infection
  marg_ci_inf_vax0 <- colMeans(ci_inf_vax0)
  marg_ci_inf_vax1 <- colMeans(ci_inf_vax1)
  ve_any_inf <- 1 - marg_ci_inf_vax1 / marg_ci_inf_vax0

  # conditional cumulative incidence symptomatic infection under no vaccine/vaccine
  ci_symp_inf_vax0 <- apply(ci_inf_vax0, 2, function(ci_inf){
  	ci_inf * p_symp__inf_vax0
  })
  ci_symp_inf_vax1 <- apply(ci_inf_vax1, 2, function(ci_inf){
  	ci_inf * p_symp__inf_vax1
  })

  # marignal parameters symptomatic infection
  marg_ci_symp_vax0 <- colMeans(ci_symp_inf_vax0)
  marg_ci_symp_vax1 <- colMeans(ci_symp_inf_vax1)
  ve_symp_inf <- 1 - marg_ci_symp_vax1 / marg_ci_symp_vax0
  
  # conditional cumulative incidence asymptomatic infection under no vaccine/vaccine
  ci_asymp_inf_vax0 <- apply(ci_inf_vax0, 2, function(ci_inf){
  	ci_inf * ( 1 - p_symp__inf_vax0 )
  })
  ci_asymp_inf_vax1_for_veai <- apply(ci_inf_vax1, 2, function(ci_inf){
  	ci_inf * ( 1 - p_symp__inf_vax1 )
  })
  ci_asymp_inf_vax1_for_venai <- apply(ci_inf_vax1, 2, function(ci_inf){
  	ci_inf * ( 1 - p_symp__inf_vax0 )
  })
  
  # marginal parameters asymptomatic infection
  marg_ci_asymp_vax0 <- colMeans(ci_asymp_inf_vax0)
  marg_ci_asymp_vax1_for_veai <- colMeans(ci_asymp_inf_vax1_for_veai)
  marg_ci_asymp_vax1_for_venai <- colMeans(ci_asymp_inf_vax1_for_venai)
  ve_ai <- 1 - marg_ci_asymp_vax1_for_veai / marg_ci_asymp_vax0
  ve_nai <- 1 - marg_ci_asymp_vax1_for_venai / marg_ci_asymp_vax0


  # principal strata estimates
  p_immune__covariates <- 1 - ci_inf_vax0
  
  p_doomed__covariates <- ci_symp_inf_vax1

	p_alwaysinf__covariates <- ci_asymp_inf_vax1_for_venai

	# p_converted__covariates <- mapply(p1 = as.list(ci_symp_inf_vax1_for_veai), p2 = as.list(ci_symp_inf_vax1_for_venai),
	# 	function(p1, p2){
	# 		p1 - ( 1 - p_symp__inf_vax0 ) * p2
	# })
	p_converted__covariates <- matrix(NA, nrow = nrow(p_alwaysinf__covariates), ncol = length(t0))
	for(j in seq_along(t0)){
		p_converted__covariates[,j]  <- ci_asymp_inf_vax1_for_veai[,j] - ( 1 - p_symp__inf_vax0 ) * ci_inf_vax1[,j]
	}

	# p_helped__covariates <- mapply(p1 = as.list(ci_symp_inf_vax0), p2 = as.list(ci_symp_inf_vax1_for_venai),
	# 	function(p1, p2){
	# 		p1 - ( 1 - p_symp__inf_vax0 ) * p2
	# 	})
	p_helped__covariates <- matrix(NA, nrow = nrow(p_alwaysinf__covariates), ncol = length(t0))
	for(j in seq_along(t0)){
		p_helped__covariates[,j]  <- ci_asymp_inf_vax0[,j] - ( 1 - p_symp__inf_vax0 ) * ci_inf_vax1[,j]
	}


	# p_helpedplus__covariates <- mapply(
	# 	p1 = as.list(p_immune__covariates), p2 = as.list(p_doomed__covariates),
	# 	p3 = as.list(p_alwaysinf__covariates), p4 = as.list(p_converted__covariates), 
	# 	p5 = as.list(p_helped__covariates),
	# 	function(p1, p2, p3, p4, p5){
	# 		1 - (p1 + p2 + p3 + p4 + p5)
	# 	}
	# )
	p_helpedplus__covariates <- matrix(NA, nrow = nrow(p_alwaysinf__covariates), ncol = length(t0))
	for(j in seq_along(t0)){
		p_helpedplus__covariates[,j] <- 1 - (
				p_immune__covariates[,j] + 
				p_doomed__covariates[,j] + 
				p_alwaysinf__covariates[,j] + 
				p_converted__covariates[,j] + 
				p_helped__covariates[,j]
		)
	}
  
  out <- list(
  	t0 = t0, ve_i = ve_any_inf, ve_s = ve_symp_inf, 
  	ve_ai = ve_ai, ve_nai = ve_nai, 
  	p_immune = colMeans(p_immune__covariates),
  	p_immune_X = p_immune__covariates,
  	p_doomed = colMeans(p_doomed__covariates),
  	p_doomed_X = p_doomed__covariates,
  	p_alwaysinf = colMeans(p_alwaysinf__covariates),
  	p_alwaysinf_X = p_alwaysinf__covariates,
  	p_converted = colMeans(p_converted__covariates),
  	p_converted_X = p_converted__covariates,
  	p_helped = colMeans(p_helped__covariates),
  	p_helped_X = p_helped__covariates,
  	p_helpedplus = colMeans(p_helpedplus__covariates),
  	p_helpedplus_X = p_helpedplus__covariates
  )
  class(out) <- "ps_ve"
  return(out)
}

print.ps_ve <- function(object, what = "ve"){
	if(what == "ve"){
		print(data.frame(
			object[c("t0", "ve_i", "ve_s", "ve_ai", "ve_nai")]
		))
	}else if(what == "ps"){
		print(data.frame(
			object[c("t0", paste0("p_", c("immune", "doomed", "alwaysinf", "converted", "helped", "helpedplus")))]
		))
	}
}

fit_ps_msm <- function(
	object,
	data,
	t0 = object$t0[1],
	which_ps_numerator = "converted",
	which_ps_denominator = c("immune", "doomed", "helped", "helpedplus", "converted", "alwaysinf"),
	msm_covariates = "age",
	msm_formula = "splines::ns(age, 3)",
	msm_family = binomial(),
	...
){
	stopifnot(t0 %in% object$t0)
	stopifnot(all(msm_covariates %in% colnames(data)))
	stopifnot(all(which_ps_numerator %in% c("immune", "doomed", "helped", "helpedplus", "converted", "alwaysinf")))

	if(which_ps_denominator == "naturally_symptomatic"){
		which_ps_denominator <- c("doomed", "helpedplus", "converted")
	}else if(which_ps_denominator == "naturally_asymptomatic"){
		which_ps_denominator <- c("alwaysinf", "helped")
	}else if(which_ps_denominator == "naturally_infected"){
		which_ps_denominator <- c("doomed", "helped", "helpedplus", "converted", "alwaysinf")
	}
	stopifnot(all(which_ps_denominator %in% c("immune", "doomed", "helped", "helpedplus", "converted", "alwaysinf")))

	msm_outcome_names_numerator <- paste0("p_", which_ps_numerator, "_X")
	msm_outcome_names_denominator <- paste0("p_", which_ps_denominator, "_X")

	t0_idx <- which(object$t0 == t0)
	
	msm_outcome_numerator <- Reduce("+", lapply(object[msm_outcome_names_numerator], function(x){
		x[,t0_idx]
	}))
	msm_outcome_denominator <- Reduce("+", lapply(object[msm_outcome_names_denominator], function(x){
		x[,t0_idx]
	}))
	msm_outcome <- msm_outcome_numerator / msm_outcome_denominator

	full_msm_formula <- paste0("msm_outcome ~ ", msm_formula)
	msm_data <- data.frame(data[,msm_covariates,drop = FALSE], msm_outcome = msm_outcome)
	
	msm_fit <- glm(
		as.formula(full_msm_formula),
		data = msm_data,
		family = msm_family
	)
	out <- list(
		msm_fit = msm_fit,
		which_ps_numerator = which_ps_numerator, 
		which_ps_denominator = which_ps_denominator
	)
	return(out)
}

```

Test it out:

```{r}
ve_fit <- ve_from_fits(
	symp_cox_fit = symp_cox_fit, 
	asymp_cox_fit = asymp_cox_fit, 
	symp_lr_fit = symp_lr_fit, 
	data = data, 
	t0 = c(25, 50, 100)
)

print(ve_fit)
print(ve_fit, what = "ps")

msm_fit <- fit_ps_msm(
	object = ve_fit,
	which_ps_numerator = "converted",
	which_ps_denominator = c("converted", "helpedplus", "doomed"),
	data = data,
	t0 = 100,
	msm_covariates = "age"
)

age_range <- seq(min(data$age), max(data$age), length = 1000)
msm_pred <- predict(msm_fit$msm_fit, newdata = data.frame(age = age_range), type = "response")
plot(msm_pred ~ age_range)
```



